---
title: "N741 Homework 8"
author: "Valerie Mac"
date: "April 12, 2017"
output:
  html_document: default
  github_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Homework 8 - DUE April 12, 2017 at 5pm

Please submit Homework 8 as a PDF to CANVAS no later than 5pm EST on April 12, 2017.

## Wisconsin Breast Cancer Data (Original)

For Homework 8 you will be working with the "Original" Wisconsin Breast Cancer dataset from the UCI Machine Learning Repository; see UCI dataset [http://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+%28Original%29](http://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+%28Original%29).

The raw data files can be downloaded from the associated Data Folder at [http://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/](http://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/). In this homework you will be working with the "breast-cancer-wisconsin.data" dataset, which is a CSV comma delimited file with NO column names in the 1st row. The datafile description and associated column file names are in the "breast-cancer-wisconsin.names" which is a simple text file [http://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/breast-cancer-wisconsin.names](http://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/breast-cancer-wisconsin.names). In this text file, as you read through it and scroll down, you'll see the following:

```
7. Attribute Information: (class attribute has been moved to last column)

   #  Attribute                     Domain
   -- -----------------------------------------
   1. Sample code number            id number
   2. Clump Thickness               1 - 10
   3. Uniformity of Cell Size       1 - 10
   4. Uniformity of Cell Shape      1 - 10
   5. Marginal Adhesion             1 - 10
   6. Single Epithelial Cell Size   1 - 10
   7. Bare Nuclei                   1 - 10
   8. Bland Chromatin               1 - 10
   9. Normal Nucleoli               1 - 10
  10. Mitoses                       1 - 10
  11. Class:                        (2 for benign, 4 for malignant)
```

So, the final datafile will have 11 columns. The dataset itself is a compilation of multiple groups of clinical cases also detailed in the breast-cancer-wisconsin.names" file [http://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/breast-cancer-wisconsin.names](http://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/breast-cancer-wisconsin.names).

The combined dataset has 699 cases (rows). However, 16 cases were missing values for the "Bare Nuclei" measurement. The R code below, processes the data, applies the names, and removes the cases with missing values. So, the final dataset created below `bcdat` will have 683 cases and 11 variables.

```{r}
# from tidyverse - use readr
# to read in the comma delimited dataset
library(readr)

# raw data does not have column names
bcdat <- read_csv("breast-cancer-wisconsin.data",
                  col_names=FALSE)

# add variable names
names(bcdat) <- c("idnum","clumpthickness","uniformcellsize",
                  "uniformcellshape","marginaladhesion",
                  "singlecellsize","barenuclei","blandchromatin",
                  "normalnucleoli","mitoses","class")

# note in column 7 "Bare Nucleoli" there are
# question marks "?" that need to be set to missing NA
library(dplyr)
bcdat <- bcdat %>%
  mutate(barenucfix = ifelse(barenuclei=="?",NA,
                              as.numeric(barenuclei)))

# keep the main 11 variables
bcdat <- bcdat %>%
  select(idnum,clumpthickness,uniformcellsize,uniformcellshape,
         marginaladhesion,singlecellsize,barenucfix,blandchromatin,  
         normalnucleoli,mitoses,class)

# keep only complete cases, n=683
bcdat <- na.omit(bcdat)
```

## Principal Components Analysis

For this Homework, please refer back to the code and exercises that Dr. Hertzberg presented during lesson 10 - specifically review towards the end of "Lesson10Part3.Rmd" see [https://github.com/vhertzb/Lesson10/blob/master/Lesson10Part3.Rmd](https://github.com/vhertzb/Lesson10/blob/master/Lesson10Part3.Rmd). During this exercise, Dr. Hertzberg introduced you to the `prcomp` procedure for performing principal components analysis. `prcomp` is part of the built-in `stats` package with base R. To learn morn type `help(prcomp)`.

In Dr. Hertzberg's example, she provided code for:

* performing the principal components analysis (pca)
* using the pca output to make a plot of the variances for each principal component (pc)
* computing the PVE (percent variance explained) and plotting the PVE
* and plotting the principal component "scores" of the cases (e.g. the "scores" plot)

I will layout the code below for running the PCA for the dataset as a whole, which will include also making a "loadings" plot for the variable "coefficients" or "loading weights" for each PC - these "loading plots" give us additional insight into (a) how the variables cluster or relate/correlate with each other or not and (b) where they fall in terms of relevance for each PC in the plot. For this dataset, we can easily get away with keeping only 2 PCs and making simplier 2D scatterplots for both the "loading plot" and "scores plot".

Use the code steps below to help you complete this homework 8 assignment.

## 1. Perform the PCA

```{r}
# use only columns 2 through 10
# you do not need the idnum, nor the class variables
pr.out <- prcomp(bcdat[,2:10], scale=TRUE)
summary(pr.out)
```

## 2. Make plots of the variance and PVE 

### Plot of the Variances of Each PC

```{r}
plot(pr.out)
```

### Plot of the PVE and Cumulative PVE of each PC

```{r}
# plots of the PVE percent variance explained
pve = 100*pr.out$sdev^2/sum(pr.out$sdev^2)
pve
plot(pve, type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="blue")
plot(cumsum(pve), type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="brown3")

```

## 3. Make a "loadings plot" of the variables

```{r}
# loadings are in the "rotation" part of the 
# pr.out list object. "rotation" is a matrix
# with a row for each variable and a column for
# each PC.
pr.out$rotation

# choose the 1st and 2nd columns for the 1st 2 PCs
# and plot these loading weights for the 9
# variables. I tweaked the limits some
# feel free to change these as needed
plot(pr.out$rotation[,1],pr.out$rotation[,2],
     xlim=c(-0.5,0.1),ylim=c(-0.5,1),
     cex=2, pch=19,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Loadings Plot for PC 1 and 2")

# add xpd=FALSE to prevent lines drawn outside plot area
par(xpd=FALSE)

# add red dashed lines for the axes at y=0 and x=0
abline(h=0, col="red")
abline(v=0, col="red")

# overlay the variable names on this loading plot
text(pr.out$rotation[,1],pr.out$rotation[,2],
     labels = rownames(pr.out$rotation),
     pos = 3)
```

## 4. Scores Plot on 1st 2 PCs

```{r}
# scores plot - use x from the pr.out list object
# plot scores on 1st 2 PCs, columns 1 and 2 of x
# color the points by the "class" variable for
# benign (class=2) or malignant (class=4)
plot(pr.out$x[,1],pr.out$x[,2], 
     col = bcdat$class,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Scores Plot on PC 1 and 2",
     sub = "Blue=Benign (class=2) and Red=Malignant (class=4)")
```

## Homework 8 Tasks

1. Rerun the PCA (steps 1-4 above) for (A) just the Benign cases and for just the (B) Malignant Cases. The code below, sets up these data subsets for you.

```{r}
# Benign cases
bcdatBenign <- bcdat %>%
  filter(class == 2)

# Malignant cases
bcdatMalignant <- bcdat %>%
  filter(class == 4)
```

_HINT: simply rename the new subsets and run the code steps above._ 

```{r}
# redo for benign ==============
bcdat <- bcdatBenign
# run steps above
```
#Steps 1-4 for Benign cases only
```{r}
## 1. Perform the PCA for Bengin Cases Only
# use only columns 2 through 10
# you do not need the idnum, nor the class variables
pr.out <- prcomp(bcdat[,2:10], scale=TRUE)
summary(pr.out)
```

## 2. Make plots of the variance and PVE 

### Plot of the Variances of Each PC

```{r}
plot(pr.out)
```

### Plot of the PVE and Cumulative PVE of each PC

```{r}
# plots of the PVE percent variance explained
pve = 100*pr.out$sdev^2/sum(pr.out$sdev^2)
pve
plot(pve, type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="blue")
plot(cumsum(pve), type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="brown3")

```

## 3. Make a "loadings plot" of the variables

```{r}
# loadings are in the "rotation" part of the 
# pr.out list object. "rotation" is a matrix
# with a row for each variable and a column for
# each PC.
pr.out$rotation

# choose the 1st and 2nd columns for the 1st 2 PCs
# and plot these loading weights for the 9
# variables. I tweaked the limits some
# feel free to change these as needed
plot(pr.out$rotation[,1],pr.out$rotation[,2],
     xlim=c(-0.5,0.1),ylim=c(-0.5,1),
     cex=2, pch=19,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Loadings Plot for PC 1 and 2")

# add xpd=FALSE to prevent lines drawn outside plot area
par(xpd=FALSE)

# add red dashed lines for the axes at y=0 and x=0
abline(h=0, col="red")
abline(v=0, col="red")

# overlay the variable names on this loading plot
text(pr.out$rotation[,1],pr.out$rotation[,2],
     labels = rownames(pr.out$rotation),
     pos = 3)
```

## 4. Scores Plot on 1st 2 PCs

```{r}
# scores plot - use x from the pr.out list object
# plot scores on 1st 2 PCs, columns 1 and 2 of x
# color the points by the "class" variable for
# benign (class=2) or malignant (class=4)
plot(pr.out$x[,1],pr.out$x[,2], 
     col = bcdat$class,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Scores Plot on PC 1 and 2",
     sub = "Blue=Benign (class=2) and Red=Malignant (class=4)")
```

```{r}
# redo for malignant ==================
bcdat <- bcdatMalignant
# run steps above
```

## 1. Perform the PCA for Malignant Cases only
```{r}
# use only columns 2 through 10
# you do not need the idnum, nor the class variables
pr.out <- prcomp(bcdat[,2:10], scale=TRUE)
summary(pr.out)
```

## 2. Make plots of the variance and PVE 

### Plot of the Variances of Each PC

```{r}
plot(pr.out)
```

### Plot of the PVE and Cumulative PVE of each PC

```{r}
# plots of the PVE percent variance explained
pve = 100*pr.out$sdev^2/sum(pr.out$sdev^2)
pve
plot(pve, type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="blue")
plot(cumsum(pve), type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="brown3")

```

## 3. Make a "loadings plot" of the variables

```{r}
# loadings are in the "rotation" part of the 
# pr.out list object. "rotation" is a matrix
# with a row for each variable and a column for
# each PC.
pr.out$rotation

# choose the 1st and 2nd columns for the 1st 2 PCs
# and plot these loading weights for the 9
# variables. I tweaked the limits some
# feel free to change these as needed
plot(pr.out$rotation[,1],pr.out$rotation[,2],
     xlim=c(-0.5,0.1),ylim=c(-0.5,1),
     cex=2, pch=19,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Loadings Plot for PC 1 and 2")

# add xpd=FALSE to prevent lines drawn outside plot area
par(xpd=FALSE)

# add red dashed lines for the axes at y=0 and x=0
abline(h=0, col="red")
abline(v=0, col="red")

# overlay the variable names on this loading plot
text(pr.out$rotation[,1],pr.out$rotation[,2],
     labels = rownames(pr.out$rotation),
     pos = 3)
```

## 4. Scores Plot on 1st 2 PCs

```{r}
# scores plot - use x from the pr.out list object
# plot scores on 1st 2 PCs, columns 1 and 2 of x
# color the points by the "class" variable for
# benign (class=2) or malignant (class=4)
plot(pr.out$x[,1],pr.out$x[,2], 
     col = bcdat$class,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Scores Plot on PC 1 and 2",
     sub = "Blue=Benign (class=2) and Red=Malignant (class=4)")
```

2. In the overall dataset, when looking at the loadings plot, which variables cluster together? which variables do not lie with that cluster?

In the overall dataset, the following variables appear to cluster together: clump thickness, Uniformity of cell size, uniformity of cell shape, marginal adhesion, single epithelial cell size, bare nuclei, bland chromatin, and normal nucleoli. 

Mitosis does not appear to cluster with the other variables.

3. How do the variable clusters seen in the loading plots for the Benign data subset and Malignant subset differ? and how are they similar if at all?

There appear to be more variable clusters (and thus more variability) in the Malignant dataset as compared to the benign dataset. Also, the variable mitosis appeared in a cluster with other variables in the malignant dataset which is quite different than its positon on the loading plot of the overall dataset and the benign dataset. 

When looking at the malignant and benign datasets, clump thickness distanced itself from the cluster it had been part of in the dataset as a whole both the malignant and benign datasets.  

Bare nuclei was a variable that also did not become apparent as a potential separate cluster until the loading plotS for the benign and malignant dataset.

Also, mitosis did not appear as a separate "cluster" in the malignant dataset as compared to the benign dataset, which may be a defining characteristic of the malignant data subset.

4. Is using 2 principal components reasonable for summarizing the variability seen in this Breast Cancer dataset with 9 measurements? Explain your reasoning for (a) the overall dataset, (b) the Benign subset and (c) the Malignant subset

Using 2 PCs is a slim approach for PCA, in general. For this dataset, using 2 PCs is reasonable for summarizing the variability, but it is not ideal. 

(a) Based upon the plot of the cumulative PVE (Percent of Variability Explained) and PCs, it appears that for the overall dataset, 2 Pcs explains the majority of the variance in the dataset, with the slope of the line dropping significantly after 2, and even more after, and so adding a third PC could be valuable. However, beyond 3 it does not appear that a fourth would help.

(b) For the benign subset, there appears to me more variance than was visually apparent in the dataset as a whole. Also, the plot of the cumulative PVE and PCs does not have as consistent of a drop off after 2 PCs. Based upon this information and the emergence of a potential 3rd cluster on the loading plot, I would prefer more than 3 PCs. 

(c) The variability appears to be even greater within the malignant subset. Due to this increased variability, I feel that more than 2 PCs is warranted, especially based upon the slope of the line on the plot of the cumulative PVE and PCs which appears to show that there are reasonable increases in cumulative PVE as principal components are added.  

5. While PCA is an unsupervised data analysis method (i.e. no "target" class information is used in the analysis), do you think the 2 PCs extracted do a good job of helping to distinguish Benign cases from Malignant cases (i.e. look back at the overall dataset Scores Plot). Explain your rationale.

I feel like 2 PCs extracted may not be sufficient for summarizing the variability within this dataset for distinguishing benign from malignant cases. According to the appearance of overall data scores plot, 2 PCs differentiates most of the cases, but when trying to differentiate such an important clinical phenotype, I feel that a more liberal use of PCs to differentiate benign from malignant would be prudent, especially due to the overlapping area of the benign and malignant cases. This perception underscored by the variability within the malignant subset, and the clustering that appears in the malignant subset loading plots.

6. Please save your RMD to a Github repository. Submit the PDF report for Homework 8 to CANVAS and include a link to your Homework 8 Github Repository.

My data repository for HW 8:
[https://github.com/valerie-mac/N741hmwk8.git](https://github.com/valerie-mac/N741hmwk8.git)
